# Red Wine Exploratory Data Analysis 

## by Javier Soto

===========================================================================================

***

## Introduction

This analysis explores the univariate, bivariate, & multivariate 
relationships between the variables in the provided tidy Red Wine data set
using RStudio. The complete dataset can be found [here.](https://www.google.com/url?q=https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityReds.csv&sa=D&ust=1539234033126000) 

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

# opts_chunk$set(tidy = FALSE, fig.width = 11)
suppressMessages(library(reshape))
suppressMessages(library(reshape2))
suppressMessages(library(gridExtra))
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
suppressMessages(library(RColorBrewer))
suppressMessages(library(GGally))
suppressMessages(library(scales))
suppressMessages(library(memisc))
suppressMessages(library(pander))
suppressMessages(library(knitr))
suppressMessages(library(xtable))
suppressMessages(library(grid))
suppressMessages(library(interp))
suppressMessages(library(utils))
suppressMessages(library(DescTools))
suppressMessages(library(lazyeval))
suppressMessages(library(Simpsons))
suppressMessages(library(corrplot))
```

***

## Red Wine Data - Load and Assessment

First step is to load the data and do some basic assessments.
This includes displaying the internal structure of the initial data frame.


```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}

# Load the Data

rw <- read.csv('wineQualityReds.csv')

# display the internal structure

str(rw)

```

***

### Red Wine Data - Transforming Quality Integer Data to a Factor

Next, programatically transformed the integer Quality data to a Factor,
and output the head and tail results for review.

```{r echo=FALSE, message=FALSE, warning=FALSE, Transforming_Quality}

#Transforming Quality from an Integer to a Factor

rw$quality <- factor(rw$quality, ordered = T)

head(rw$quality)
tail(rw$quality)

```

***

### Red Wine Data - Create New Rating Factor
Generated a new Factored Variable named 'Rating,'
and output the head and tail results for review.

```{r echo=FALSE, message=FALSE, warning=FALSE, Create_New_Rating_Factor}

#Creating a new Factored Variable called 'Rating'

rw$rating <- ifelse(rw$quality < 5, 'bad', ifelse(
  rw$quality < 7, 'average', 'good'))

rw$rating <- ordered(rw$rating,
                       levels = c('bad', 'average', 'good'))

head(rw$rating)
tail(rw$rating)

```


***

### Red Wine Data - Display New Internal Structure

Display the internal structure after creating the new Factored Variable: Rating.


```{r echo=FALSE, message=FALSE, warning=FALSE, Display_New_Internal_Structure}

# display the internal structure after creating the new Factored Variable: Rating

str(rw[2:14])

```

***

### Red Wine Data - Load and Assessment

In addition, listed out the column (variable) names of interest for reference.
Superfluous columns were excluded.

```{r echo=FALSE, message=FALSE, warning=FALSE, Column_Names}

pandoc.list(names(rw[2:14]))

```

***

### Red Wine Data - Display Head of Selected Columns

Output of the head for selected columns.

```{r echo=FALSE, Summary_Head}

head(rw[10:14], 5)

```

***

### Red Wine Data - Display Tail of Selected Columns

Output of the tail for selected columns.

```{r echo=FALSE, Summary_Tail}

tail(rw[10:14], 5)

```


***

### Red Wine Data - Display Summary Statistics

Output of the Summary Statistics together with the Counts 
for the Factored Variables:

```{r echo=FALSE, message=FALSE, warning=FALSE, Summary_Stats_01}

# display the summary statistics

table1 <- summary(rw)

tt3 <- ttheme_minimal(
  core=list(bg_params = list(fill = blues9[1:4], col=NA),
            fg_params=list(fontface=3)),
  colhead=list(fg_params=list(col="navyblue", fontface=4L)),
  rowhead=list(fg_params=list(col="orange", fontface=3L)))

grid.arrange(
  tableGrob(table1[1:6, 2:5], theme = tt3),
  tableGrob(table1[1:6, 6:9], theme = tt3),
  nrow=2)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Summary_Stats_02}

# display the summary statistics

grid.arrange(
  tableGrob(table1[1:6, 10:12], theme = tt3),
  tableGrob(table1[1:6, 13:14], theme = tt3),
  nrow=2)

```

***

## Univariate Plots Section

First lets plot the distribution of each variable to get a sense of the data.
Now, lets remove some outliers and explore these individual univariate plots 
in detail!

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_Arranged}

# create plots to arrange 

p0 <- ggplot(data = rw, aes(x = alcohol), binwidth = 0.01) +
              geom_histogram(color = I('Black'), fill = I('#1518c1')) 

p1 <- ggplot(data = rw, aes(x = residual.sugar), binwidth = 1) +
              geom_histogram(color = I('Black'), fill = I('#1518c1'))

p2 <- ggplot(data = rw, aes(x = pH), binwidth = 1) +
              geom_histogram(color = I('black'), fill = I('#1518c1'))

p3 <- ggplot(data = rw, aes(x = density), binwidth = 1) +
              geom_histogram(color = I('black'), fill = I('#1518c1'))

p4 <- ggplot(data = rw, aes(x = chlorides), binwidth = 1) +
              geom_histogram(color = I('black'), fill = I('#1518c1'))

p5 <- ggplot(data = rw, aes(x = free.sulfur.dioxide), binwidth = 1) +
              geom_histogram(color = I('black'), fill = I('#1518c1'))

p6 <- ggplot(data = rw, aes(x = total.sulfur.dioxide), binwidth = 1) +
              geom_histogram(color = I('black'), fill = I('#1518c1'))

p7 <- ggplot(data = rw, aes(x = sulphates), binwidth = 1) +
              geom_histogram(color = I('black'), fill = I('#1518c1'))


p8 <- ggplot(data = rw, aes(x = citric.acid), binwidth = 1) +
              geom_histogram(color = I('black'), fill = I('#1518c1'))


# arrange plots in 3 columns
suppressMessages(grid.arrange(p0, p1, p2, p3, p4, p5, p6, p7, p8, ncol = 3))
              
```
          

***

### Univariate Plots - pH and Density Normally Distributed

Both distribution plots for pH and Density appear to be normally distributed.

The mean typically is equal to the median.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_Normal_Density}

 ggplot(data = rw, aes(x = density), binwidth = 1) +
              geom_histogram(color = I('black'), fill = I('#1518c1')) +
ggtitle("Density Histogram between 0.9930 and 1.0000

Mean and Median equal .997          
          
") + 
  coord_cartesian(xlim=c(0.993,1.00))

#grid.arrange(p9, p10, ncol =2)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_Normal_pH}

ggplot(data = rw, aes(x = pH), binwidth = 1) +
              geom_histogram(color = I('black'), fill = I('#1518c1')) +
ggtitle("pH Histogram between 3.00 and 3.60

Mean and Median equal 3.31

") + 
  coord_cartesian(xlim=c(3.00,3.6))

```

***

### Univariate Plots - Right Skewed Distributions

The distribution plots for alcohol, residual sugar, chlorides, free sulfur
dioxide, total sulfur dioxide, and sulphates appear to be rightly skewed. 

For a right skewed distribution, the mean is typically greater than the median. 
Also, the tail of the distribution on the right hand (positive) side is longer 
than on the left hand side.


```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_Right_Alcohol}

ggplot(data = rw, aes(x = alcohol), binwidth = 1) +
              geom_histogram(color = I('black'), fill = I('#1518c1')) +
ggtitle("Alcohol Histogram between 9.00 and 13.00

Mean 10.42 is greater than median 10.20
") + 
  coord_cartesian(xlim=c(9.5,13.0))

```


```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_Right_Residual_Sugar}

ggplot(data = rw, aes(x = residual.sugar), binwidth = 1) +
              geom_histogram(color = I('black'), fill = I('#1518c1')) +
ggtitle("Residual Sugar Histogram between 1.50 and 7.00

Mean 2.539 is greater than median 2.20
") + 
  coord_cartesian(xlim=c(1.5,7))

```


```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_Right_Chlorides}

ggplot(data = rw, aes(x = chlorides), binwidth = 1) +
              geom_histogram(color = I('black'), fill = I('#1518c1')) +
ggtitle("Chlorides Histogram between 0.05 and 0.30

Mean 0.087 is greater than median 0.079
") + 
  coord_cartesian(xlim=c(0.05,0.3))

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_Right_Free_Sulfur}

ggplot(data = rw, aes(x = free.sulfur.dioxide), binwidth = 1) +
              geom_histogram(color = I('black'), fill = I('#1518c1')) +
ggtitle("Free Sulfur Dioxide Histogram between 6.00 and 45.00

Mean 15.87 is greater than median 14.00
") + 
  coord_cartesian(xlim=c(6.00,45.00))

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_Right_Total_Sulfur}

ggplot(data = rw, aes(x = total.sulfur.dioxide), binwidth = 1) +
              geom_histogram(color = I('black'), fill = I('#1518c1')) +
ggtitle("Total Sulfur Dioxide Histogram between 5.00 and 160.00

Mean 46.47 is greater than median 38.00
") + 
  coord_cartesian(xlim=c(5.00,160.00))

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_Right_Total_Sulphates}

ggplot(data = rw, aes(x = sulphates), binwidth = 1) +
              geom_histogram(color = I('black'), fill = I('#1518c1')) +
ggtitle("Sulphates Histogram between 0.50 and 1.25

Mean 0.6581 is greater than median 0.6200
") + 
  coord_cartesian(xlim=c(.5,1.25))

```

***

### Univariate Plots - Citric Acid is Bimodally Distributed

The distribution plot for Citric Acid appears to be Bimodally Distributed.

In statistics, a bimodal distribution appears as distinct peaks.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_Bimodal_Citric_Acid}

ggplot(data = rw, aes(x = citric.acid), binwidth = 1) +
              geom_histogram(color = I('black'), fill = I('#1518c1')) +
ggtitle("Sulphates Histogram between 0.10 and 0.75

Citric Acid is Bimodally Distributed
") + 
  coord_cartesian(xlim=c(.1, 0.75))

```

***

### Univariate Plots - Quality and Ratings

Most of the wines in the dataset are average quality wines. I need to determine
the main variables that are driving these results.

```{r echo=FALSE, Univariate_Plots_Rating_Quality}

p13 <- ggplot(data = rw, aes(x = quality)) +
  geom_bar(width = .5, color = 'black',fill = I('#1518c1'))

p14 <- ggplot(data = rw, aes(x = rating)) +
  geom_bar(width = .25, color = 'black',fill = I('#1518c1'))

grid.arrange(p13, p14, ncol =2)

```

***

## Analysis of the Univariate Plots

### Structure of the dataset?

The Red Wine Dataset has 1599 rows and 13 columns. The categorical variables 
are 'quality' and 'rating', and the remainder are numerical variables that 
indicate the physical and chemical properties of the wine.

It was observed that most of the wines are categorized as 'average' quality 
with a few 'bad' and 'good' wines, respectively, in the tals. Also, all 
variables showed outliers. This suggests that the data is of poor predictive 
value.

### Main Feature

My main point of interest in this dataset is the average 'rating', and why so 
many wines fell into that bucket. I would like to determine which factors 
determine the quality of a wine and what factors are critical for a good wine 
versus a bad wine.

### Other Features

The variables related to acidity (fixed, volatile, citric.acid and pH) might 
explain some of the variance. The different acid concentrations might 
affect the taste of the wine. Also,residual sugar determines how sweet a wine
is, so it also may influence the taste.

### New Variable Created

A rating variable was created.

### Unusual Distributions

Citric.acid stood out from the other distributions. It had outliers and what 
appeared to be a bimodal distribution. Also, most of the variables were rightly 
skewed. Outliers were removed to confirm that the observed bimodal and skewed
distributions were valid, and not the result of a few outliers.

***

## Bivariate Plots Section

### First step is to generate a correlation table, for all of the relevant variables in the provided red wine data set, to better understand the relationships between them.


```{r echo=FALSE, Bivariate_Plots}

# to calculate the correlation matrix transform the data

c <- cor(
  rw %>%
    
    # remove unwanted columns
    
    dplyr::select(-X) %>%
    dplyr::select(-rating) %>%
    mutate(
      
      # translate quality to a number
      
        quality = as.numeric(quality)
      )
  )

# highlight correlations greater than 0.3 and less than -0.3 

emphasize.strong.cells(which(abs(c) > .3 & c != 1, arr.ind = TRUE)) 


# display and format correlation table

set.alignment('right', row.names = 'left')
# pandoc.table(c[1:12, ], style = "grid", round = 2, plain.ascii = TRUE)
pandoc.table(c[1:12, ], style = "grid", round = 2)
```

***

### Second step is to create and leverage a function to programmatically generate summary tables alongside selected bivariate plots between these variables for analysis of the selected correlations.

```{r echo = FALSE, message = FALSE, warning = FALSE, Summary_Plot_Function}

# function to output summaries

summary_plot <- function(data, x, y) {
  pandoc.table(data %>%
  group_by_(x) %>%
  summarise_(
    mean = interp(~mean(v), v=as.name(y)),
    median = interp(~median(v), v=as.name(y))
    ), caption = paste("Summaries for", y, "grouped by", x))
}

```

***

### Residual Sugar and Quality

Surprisingly residual sugar and quality had a very weak positive quality 
correlation of only 0.01. This suggests residual sugar impact on quality is 
negligible.

```{r echo = FALSE, message = FALSE, warning = FALSE, Plot_Residual_Sugar_Quality}

# Residual Sugar and quality have a very weak 0.01 positive correlation.

 ggplot(data=rw,
        aes(y = residual.sugar, x = quality)) +
    geom_boxplot(aes(color = rating)) +
    theme(axis.text=element_text(size = 12)) + 
    scale_color_brewer(type = 'qual', palette=6)

summary_plot(rw, "rating",  "residual.sugar")

```

***

### Volatile Acidity and Quality

Volatile acidity and quality have a -0.39 moderate negative correlation. 
This suggests red wine quality decreases as volatile acidity increases.

```{r echo = FALSE, message = FALSE, warning = FALSE, Plot_Volatile_Acid_Quality_1}

# Volatile acid and quality have a -0.39 moderate negative correlation, which implies red wine quality decreases as volatile acidity increases.

 ggplot(data=rw,
        aes(y = volatile.acidity, x = quality)) +
    geom_boxplot(aes(color = rating)) +
    theme(axis.text=element_text(size = 12)) + 
    scale_color_brewer(type = 'qual', palette=6)

summary_plot(rw, "rating",  "volatile.acidity")

```

***

### Fixed Acidity and Quality

However, fixed acidity and quality has a weak positive correlation of 0.11. 
Fixed Acidity may have some minimal impact on wine quality.

```{r echo = FALSE, message = FALSE, warning = FALSE, Plot_Fixed_Acid_Quality}

# Fixed acidity and quality have a weak correlation

 ggplot(data=rw,
        aes(y = fixed.acidity, x = quality)) +
    geom_boxplot(aes(color = rating)) +
    theme(axis.text=element_text(size = 12)) + 
    scale_color_brewer(type = 'qual', palette=6)

summary_plot(rw, "rating",  "fixed.acidity")

```

***

### Quality and Sulphates & Quality and Citric acid.

Furthermore, there are weak positive correlations for both (1) quality and 
sulphates at .25 and (2) quality and citric acid at .23. Also, ratings trends 
in the same direction for both. This suggests that better wines may have a 
stronger concentration of sulphates; and also, better wines may have higher 
citric acid.

```{r echo = FALSE, message = FALSE, warning = FALSE, Plot_Sulphates_Quality}

# There are weak positive correlations for both 1) quality and sulphates and 2) quality and citric
# acid.  Ratings trends in the same direction for both sulphates and citric acid.   

  ggplot(data=rw,
        aes(y = sulphates, x = quality)) +
    geom_boxplot(aes(color = rating)) +
    theme(axis.text=element_text(size = 12)) + 
    scale_color_brewer(type = 'qual', palette=6)

summary_plot(rw, "rating",  "sulphates")

```

```{r echo = FALSE, message = FALSE, warning = FALSE, Plot_Citric_Acid_Quality}

  ggplot(data=rw,
        aes(y = citric.acid, x = quality)) +
    geom_boxplot(aes(color = rating)) +
    theme(axis.text=element_text(size = 12)) + 
    scale_color_brewer(type = 'qual', palette=6)

summary_plot(rw, "rating",  "citric.acid")

```

***

### Citric Acid and Fixed Acidity

As expected, citric acid and fixed acidity have a strong positive correlation 
of 0.67

```{r echo = FALSE, message = FALSE, warning = FALSE, Plot_Fixed_Citric}
# Citric acid and fixed acidity have a strong positive correlation of 0.67;
    
  ggplot(data = subset(rw, rw$citric.acid > 0),
        aes(y=fixed.acidity, x = citric.acid)) +
  geom_point(alpha = 0.25, color="#56B4E9") +
  geom_smooth(color="darkblue")

```

***

### Moderate Negative Correlation between Volatile Acidity and Citric Acid

Also, you can see a moderate negative correlation between volatile acidity and 
citric acid where the volatile acidity y values scale with the square root 
function. This is not so surprising given that we observed that red wine quality 
decreases as volatile acidity increases, and better wines have higher citric acid. 

```{r echo = FALSE, message = FALSE, warning = FALSE, plot_fixac_citric_2}

#The moderate negative correlation volatile acidity and citric acid where the volatile acidity y values scale with the sqaure root function.
  
  ggplot(data=subset(rw, rw$citric.acid > 0),
        aes(y=volatile.acidity, x = citric.acid)) +
  geom_point(alpha = 0.25, color="#56B4E9") +
  scale_y_sqrt() +
  scale_x_continuous() +
  geom_smooth(color="darkblue")
  
```

***

### The Simpson function against citric acid and fixed acidity detects 7 clusters.

Running the Simpson function against citric acid and fixed acidity detected 7 
clusters. Only two clusters correlated in the same direction as the group. The 
overall trend for the subgroups reversed or disappeared when the subgroups were 
combined.

```{r echo = FALSE, message = FALSE, warning = FALSE, include = FALSE, simp_citric.acid_fixed.acidity_1}

#Running the Simpson function against citric acid and fixed acidity detects 7 clusters.  Only two
#clusters correlate in the same direction as the group.  The overall trend for the subgroups reverses
#or disappears when the subgroups are combined.

simpsons <- Simpsons(citric.acid, fixed.acidity, data = rw)

```

```{r echo = FALSE, message = FALSE, warning = FALSE, simp_citric.acid_fixed.acidity_2}

plot(simpsons)
```

***

## Bivariate Analysis

### Some of the relationships observed

Volatile acidity and quality have a -0.39 moderate negative correlation. This 
suggests red wine quality decreases as volatile acidity increases. However, 
fixed acidity and quality have a weak positive correlation. Fixed Acidity has 
almost no impact on wine quality.  Furthermore, there are weak positive 
correlations for both (1) quality and sulphates and (2) quality and citric acid. 
Also, ratings trends in the same direction for both. This suggests that better 
wines have a stronger concentration of sulphates. Also, better wines have higher 
citric acid. As expected, citric acid and fixed acidity have a strong positive 
correlation of 0.67. Also, you can see a moderate negative correlation between 
volatile acidity and citric acid where the volatile acidity y values scale with 
the square root function. This is not so surprising given that we observed that 
red wine quality decreases as volatile acidity increases, and better wines have 
higher citric acid.

### Interesting Relationships 

[Simpson's Paradox](https://en.wikipedia.org/wiki/Simpson%27s_paradox) is a 
"phenomenon in probability and statistics, in which a trend appears in several 
different groups of data but disappears or reverses when these groups are 
combined." Running the Simpson function against citric acid and fixed acidity 
detected 7 clusters. Only two clusters correlated in the same direction as the 
group. The overall trend for the subgroups reversed or disappeared when the 
subgroups were combined.

### Strongest Relationship

The strongest relationship this analysis focused on was that between citric acid 
and fixed volatility.

***

## Multivariate Plots Section

Lets use multivariate plots to answer some questions that came to light from the 
above bivariate plot analysis and to look for other relationships in the data.

***

### pH and Acid Concentration

pH measures acid concentration using a log scale. Therefore, there are stronger 
correlations between pH and the log of the acid concentrations. We use a linear 
model to investigate how much of the variance in pH is explained by citric acid, 
fixed acidity, and volatile acidity. With R-squared equal to 0.4876, it seems 
that the three acidity variables can only explain about half the variance in pH. 
This suggests there are other more relevant variables that affect acidity.

```{r echo = FALSE, message = FALSE, warning = FALSE, pH_Linear_Model_1}

acidLM <- lm(pH ~ 
                I(log10(citric.acid)) +
                I(log10(volatile.acidity)) +
                I(log10(fixed.acidity)),
                    data = subset(rw,citric.acid > 0))
summary(acidLM)
pH_df = data.frame(rw$quality )
pH_df $predictions <- predict(acidLM, rw)
pH_df $error <- (pH_df$predictions - rw$pH)/rw$pH

```

***

### Alcohol Content and Sulphates

The plots below indicate that for wines with high alcohol content, having a 
higher concentration of sulphates produces better wines.

```{r echo = FALSE, message = FALSE, warning = FALSE, alc_sulphates1}

ggplot(data = rw,
       aes(y = sulphates, x = alcohol,
           color = quality)) +
  geom_point() +
  scale_y_continuous(limits=c(0.3,1.5)) +
  scale_color_brewer(type='div')
```

```{r echo = FALSE, message = FALSE, warning = FALSE, alc_sulphates2}

ggplot(data = rw,
       aes(y = sulphates, x = alcohol,
           color = quality)) +
  geom_point() +
  scale_y_continuous(limits=c(0.3,1.5)) +
  facet_wrap(~rating) +
  scale_color_brewer(type='div')

```

Also, the inverse seems to hold for acidity. For example, having less volatile 
acidity on higher concentrations of alcohol seems to produce better wines.

```{r echo = FALSE, message = FALSE, warning = FALSE, Vola_Alc}

ggplot(data = rw,
       aes(y = volatile.acidity, x = alcohol,
           color = quality)) +
  geom_point() +
  scale_color_brewer(type='div')

```

```{r echo = FALSE, message = FALSE, warning = FALSE, Vola_pH2}

ggplot(data = rw,
       aes(y = volatile.acidity, x = alcohol,
           color = quality)) +
  geom_point() +
  facet_wrap(~rating) +
  scale_color_brewer(type='div')

```

### Key Variables Linear Models

A few selected key variables (alcohol, sulphates, and acidity) were used
to generate some linear models for comparison. The pH variable was excluded 
to avoid issues with perfect multicollinearity (also collinearity). 
Multicollinearity is a "phenomenon in which one predictor variable in a 
multiple regression model can be linearly predicted from the others with a 
substantial degree of accuracy."  The results were disappointing, with 
[R-squared](https://en.wikipedia.org/wiki/Coefficient_of_determination), the 
coefficient of determination statistic used to measure the proportion of the 
variance in the dependent variable that is explained by the independent 
variable(s), coming in low (highest reported R-square figure was .348)

```{r echo = FALSE, message = FALSE, warning = FALSE, Lm_Acidity1}

set.seed(12345)
training_data <- sample_frac(rw, .6)
test_data <- rw[ !rw$X %in% training_data$X, ]
m1 <- lm(as.numeric(quality) ~ alcohol, data = training_data)
m2 <- update(m1, ~ . + sulphates)
m3 <- update(m2, ~ . + volatile.acidity)
m4 <- update(m3, ~ . + citric.acid)
m5 <- update(m4, ~ . + fixed.acidity)
m6 <- update(m2, ~ . + pH)
mtable(m1,m2,m3,m4,m5,m6)
df <- data.frame(
  test_data$quality,
  predict(m5, test_data) - as.numeric(test_data$quality)
)
names(df) <- c("quality", "error")
```


```{r echo = FALSE, message = FALSE, warning = FALSE,  Lm_Acidity2}

ggplot(data=df, aes(x=quality,y=error)) +
  geom_point(color = "Blue")
  
```

***

## Multivariate Analysis

### Observed Relationships

We used multivariate plots to answer some questions that came to light from the 
earlier bivariate plot analysis and to look for other relationships in the data. 
We know that pH measures acid concentration using a log scale; therefore, 
there are stronger correlations between pH and the log of the acid concentrations. 
We used a linear model to investigate how much of the variance in pH is explained 
by citric acid, fixed acidity, and volatile acidity. With R-squared equal to 0.4876, 
it seems that the three acidity variables can only explain about half the variance 
in pH. This suggests there are other more relevant variables that affect acidity.

### Interesting or Surprising Interactions between Features

It was shown that wines with high alcohol content, having a 
higher concentration of sulphates produces better wines. 
Also, the inverse seems to hold for acidity. For example, having less volatile 
acidity on higher concentrations of alcohol seems to produce better wines.

### Linear Models

We also used some key variables (alcohol, sulphates, and acidity) to generate 
a few linear models for comparison. The pH variable was excluded to avoid issues 
with perfect multicollinearity (also collinearity). 
[Multicollinearity](https://en.wikipedia.org/wiki/Multicollinearity) 
is a "phenomenon in which one predictor variable in a multiple regression model 
can be linearly predicted from the others with a substantial degree of accuracy."
Also the low R-squared scores suggest that there are missing variables that can be 
better used to  predict quality.

***

## Final Plots and Summary

***

### Plot One

```{r echo=FALSE, Plot_One}

corrplot(c, order = "hclust", addrect = 2)

```

***

### Description One

For my initial analysis, originally a table was used to display all the 
correlations for the red wine data set; but plotting the correlation matrix 
makes it even easier to identify both positive and negative correlations 
greater than an absolute value of 0.2.

***

### Plot Two

```{r echo = FALSE, message = FALSE, warning = FALSE, Plot_Two_b}

ggplot(data = rw,
       aes(y = sulphates, x = alcohol,
           color = quality)) +
  geom_point(alpha = .75, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_y_continuous(limits=c(0.3,1.5)) +
  ylab("Potassium Sulphate (g/dm3)") +
  xlab("Alcohol Percentages") +
  theme(axis.text=element_text(size = 12)) + 
  scale_color_brewer(type='div',
                   guide=guide_legend(title='Red Wine 
Quality Levels')) +
  ggtitle("Alcohol and Sulphates over Red Wine Quality Levels")

```

***

### Description Two

This plot shows that the best quality wines have high concentrations fof both
alcohol and sulphates. This implies that high alcohol contents and  high 
sulphate concentrations together seem to produce better wines. This is something 
to definitely look for when shopping for red wines.

***

### Plot Three

```{r echo = FALSE, message = FALSE, warning = FALSE, Plot_Three}

df <- data.frame(
  test_data$quality,
  predict(m5, test_data) - as.numeric(test_data$quality)
)
names(df) <- c("quality", "error")

ggplot(data=df, aes(x=quality, y=error, color = quality)) +
  geom_point() +
  theme(axis.text=element_text(size = 12)) + 
  scale_color_brewer(type = 'qual', guide=guide_legend(title='Red Wine 
Quality Levels'), 
                     palette=6) +
  ylab("Linear Model Errors") +
  xlab("Quality Levels") +
  ggtitle("Linear Model Errors over Expected Quality")

```

### Description Three

The linear model with the highest R-squared value could only explain approximately
35% of the variance in quality. This plot suggests that there are missing variables 
needed to better predict quality wines.

# Reflection

This analysis explored the univariate, bivariate, & multivariate 
relationships between the variables in the provided tidy Red Wine data set
The data set contained information on the chemical properties of a selection of 
red wines. Also, a new Factored Variable named 'Rating' was added.

The first step was to do a univariate analysis of all the variables. A series of plots
showed which variables were normally distributed and skewed. Focusing on the plots
for quality and ratings showed that most wines in the dataset were of average quality.
Why?

Continuing with the bivariate analysis it was shown that better wines have
a stronger concentration of sulphates, together with higher counts of citric acid.
Then following through with a multivariate analysis it was shown that wines with 
high alcohol content and having a higher concentration of sulphates produced better 
wines. Also, the inverse seemed to hold for acidity. For example, having less volatile 
acidity on higher concentrations of alcohol seemed to produce better wines. However, 
the generated liniear models using the alcohol, sulphates, and acidity variables
did not well explain the variance in quality.

Future studies, should include larger data sets and more variables. For example, 
a [winefolly.com](https://winefolly.com/review/understanding-acidity-in-wine/) 
article speaks to the importance of differences in tastes between unoaked versus 
oaked wines and the role the aging process plays in a wine’s malic acid 
conversion to lactic acid. Also, the data set can include categorized wine critic reviews and ratings, 
to determine if relying on such critical [ratings](https://www.winespectator.com/wineratings) 
can truly lead to selecting better tasting wines.   

***

## Sources of Inspiration

**https://stackoverflow.com/questions/7458796/how-to-suppress-qplots-binwidth-warning-inside-a-function**

**https://cran.r-project.org/web/packages/gridExtra/vignettes/tableGrob.html**

**https://rapporter.github.io/pander/pandoc_table.html**

**https://github.com/pcasaretto/udacity-eda-project/blob/master/wine.Rmd**

**http://www.sthda.com/english/wiki/ggplot2-colors-how-to-change-colors-automatically-and-manually**

**https://rpubs.com/jasonmedina/219996**

**https://github.com/pcasaretto/udacity-eda-project/blob/master/wine.Rmd**

**https://ggplot2.tidyverse.org**